---
title: "Getting Started with orchestr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with orchestr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Installation

```{r install}
# install.packages("remotes")
remotes::install_github("ian-flores/orchestr")
```

## API Key Setup

orchestr uses [ellmer](https://github.com/tidyverse/ellmer) for LLM access.
Set your provider's API key before running any examples:

```{r api-key}
# For Anthropic (Claude)
Sys.setenv(ANTHROPIC_API_KEY = "your-key-here")

# For OpenAI
Sys.setenv(OPENAI_API_KEY = "your-key-here")
```

See ellmer's documentation for all supported providers.

## Using Different Providers

orchestr works with any ellmer chat backend. Pass the appropriate `chat_*()`
constructor to `agent()`:

```{r providers}
library(orchestr)
library(ellmer)

# OpenAI
agent("analyst", chat = chat_openai(model = "gpt-4o"))

# Google Gemini
agent("analyst", chat = chat_google_gemini(model = "gemini-1.5-pro"))

# Claude via AWS Bedrock
agent("analyst", chat = chat_aws_bedrock(
  model = "anthropic.claude-3-5-sonnet-20241022-v2:0"
))

# GPT-4o via Azure OpenAI
agent("analyst", chat = chat_azure_openai(
  endpoint = "https://my-resource.openai.azure.com",
  deployment_id = "gpt-4o"
))
```

All graph types (`react_graph()`, `pipeline_graph()`, `supervisor_graph()`)
work identically regardless of which provider backs the agent.

## Your First Agent

An `Agent` wraps an ellmer `Chat` object. The `agent()` constructor is the
recommended way to create one.

```{r single-agent}
library(orchestr)
library(ellmer)

analyst <- agent("analyst", chat = chat_anthropic(
  system_prompt = "You are a data analyst. Analyze data and provide insights."
))

# Single-turn conversation
response <- analyst$invoke("Describe the key features of the iris dataset.")
cat(response)
```

## Adding Tools

Agents become more powerful when you give them tools. ellmer's `Chat` class
handles tool call loops internally -- when an agent has registered tools, they
are executed automatically during `$chat()`.

```{r tools}
summary_tool <- tool(
  function(dataset_name) {
    data <- get(dataset_name, envir = asNamespace("datasets"))
    paste(capture.output(summary(data)), collapse = "\n")
  },
  "Get a summary of a built-in R dataset.",
  arguments = list(
    dataset_name = type_string("Name of a dataset in the datasets package")
  )
)

analyst <- agent("analyst",
  chat = chat_anthropic(
    system_prompt = "You are a data analyst. Use your tools to examine data."
  ),
  tools = list(summary_tool)
)

response <- analyst$invoke("Summarize the mtcars dataset.")
cat(response)
```

## Single-Agent Graph with `react_graph()`

The `react_graph()` function wraps a single agent with state management and
checkpointing. This is the simplest way to run an agent inside a graph.

```{r react}
analyst <- agent("analyst", chat = chat_anthropic(
  system_prompt = "You are a data analyst. Analyze data and provide insights."
))

graph <- react_graph(analyst)
result <- graph$invoke(list(messages = list(
  "What are the key relationships in the mtcars dataset?"
)))
```

Use `verbose = TRUE` when compiling to see execution flow. With the
convenience functions, pass `verbose` to `$invoke()`:

```{r react-verbose}
result <- graph$invoke(
  list(messages = list("Describe the distribution of mpg in mtcars.")),
  verbose = TRUE
)
```

## Agent Pipeline with `pipeline_graph()`

Chain multiple agents in sequence. Each agent processes the state and passes
it to the next. One LLM call per agent.

```{r pipeline}
profiler <- agent("profiler", chat = chat_anthropic(
  system_prompt = "Profile datasets: describe columns, types, missing values, distributions."
))

analyst <- agent("analyst", chat = chat_anthropic(
  system_prompt = "Given a data profile, identify patterns, correlations, and anomalies."
))

pipeline <- pipeline_graph(profiler, analyst)
result <- pipeline$invoke(list(messages = list(
  "Analyze the mtcars dataset focusing on fuel efficiency factors."
)))
```

## Next Steps

- **[Multi-Agent Workflows](multi-agent.html)** -- pipelines, supervisor routing, and visualization
- **[Secure Execution](securer.html)** -- sandboxed code execution with securer
